---
title: "ANOVA StatisticsAnalysis"
output: html_notebook
---

Loading the dataset. Original dataset in an excel file. 

```{r}
library(readxl)
coop.df <- read_excel("C:/AnovaStatisticalAnalysis.xlsx")
```

Summary Statistics

```{r}
library(psych)
tapply(coop.df$Cooperacao, coop.df$Modalidade, describe)
tapply(coop.df$Cooperacao, coop.df$Modalidade, summary)

```
BoxPlot

```{r}
require(ggplot2)  
graph_box <- ggplot(coop.df, aes(x = Modalidade, y = Cooperacao, fill = Modalidade)) + geom_boxplot() +
  theme_classic() + scale_fill_brewer() + stat_summary(fun.y=mean, geom="point", shape=10, size=3)
graph_box

```

Histogram to check the distribution

```{r}
ggplot(coop.df,aes(x=Cooperacao))+geom_histogram(position="identity",alpha=0.5,binwidth=0.55) + facet_grid(~Modalidade)+theme_bw() + xlab("Cooperação") + ylab("Frequency") + ggtitle("Histogram") + theme(plot.title = element_text(hjust = 0.5))
```
Checking the QQ plots for normal distribution

```{r}
qqnorm(coop.df$Cooperacao[coop.df$Modalidade=="Futebol"], main="Football Normality QQ Plot")
qqline(coop.df$Cooperacao[coop.df$Modalidade=="Futebol"])

qqnorm(coop.df$Cooperacao[coop.df$Modalidade=="Andebol"], main="Handball Normality QQ Plot")
qqline(coop.df$Cooperacao[coop.df$Modalidade=="Andebol"])

qqnorm(coop.df$Cooperacao[coop.df$Modalidade=="Rugby"], main="Rubgy Normality QQ Plot")
qqline(coop.df$Cooperacao[coop.df$Modalidade=="Rugby"])

```

Mean Plot

```{r}
library("gplots")

plotmeans(Cooperacao ~ Modalidade, data = coop.df,
          xlab = "Sports", ylab = "Cooperation",
          main="Mean Plot with 95% CI") 
```


Shapiro Wilk test for Normality

```{r}
shapiro.test(coop.df$Cooperacao)
tapply(coop.df$Cooperacao, coop.df$Modalidade, shapiro.test)
```
Levene test for assessing the equality of variances

```{r}
library(car)
groups <- factor(rep(letters[1:3], each = 37))
leveneTest(coop.df$Cooperacao, groups)
```
With all the assumptions passed, let's apply ANOVA

```{r}
my_aov <- aov(Cooperacao ~ Modalidade, data=coop.df)
summary(my_aov)
```

Tukey HSD to check which means are significantly different
```{r}
tuk <- TukeyHSD(my_aov)
tuk
```
Plotting the different pairs of means

```{r}
coop.hsd = data.frame(TukeyHSD(my_aov, which = "Modalidade")$Modalidade)

coop.hsd$Comparison = row.names(coop.hsd)


ggplot(coop.hsd, aes(Comparison, y = diff, ymin = lwr, ymax = upr)) +
  geom_pointrange() + ylab("Difference in Mean Cooperation Levels by Sport") +
  coord_flip()
```

