---
title: "Principal Component Analysis of Portuguese National Expenses"
---

Analysis of the dataset of national expenses in percentage

Importing dataset from excel and visualizing. Data originally from PORDATA (http://www.pordata.pt)
```{r}
library(readxl)

df <- read_excel("C:/PCA_Portuguese National Expenses")

row.names(df) <- df$Anos
head(df)

```

### Summary Univariate Statistics

```{r}
library(psych)
describe(df[,-1])

```

Histograms of all the variables to check distribution

```{r}

library(reshape2)

multi.hist(df[,2:6], main=" ")
multi.hist(df[,7:11], main=" ")

```
Boxplot with all the variables to check their differences


```{r}
library(ggplot2)
ggplot(stack(df[,-1]), aes(x = ind, y = values)) + geom_boxplot()
```



### Bivariate Statistics

Calculating the correlation matrix.

```{r}
cor.mat <- round(cor(df[,-1]),2)
head(cor.mat)

```

```{r}
library("corrplot")
corrplot(cor.mat, type="lower", tl.col="black", tl.srt=45)

```

### PCA

Applying PCA with correlation matrix and plotting the individuals factor map and variables correlation circle

```{r}
library(FactoMineR)

#Standartized data - Using correlation matrix
res.pca <- PCA(df[,-1], scale.unit=TRUE, ncp=5, graph=T)
```

Checking the eigenvalues and getting the scree plot

```{r}
res.pca$eig

barplot(res.pca$eig[, 2], names.arg=1:nrow(res.pca$eig), 
        main = "Scree Plot",
        xlab = "Principal Components",
        ylab = "Percentage of inertia",
        col ="steelblue")
lines(x = 1:nrow(res.pca$eig), res.pca$eig[, 2], 
      type="b", pch=19, col = "red")
```
#### Variables PCA Analysis

CTA - (Absolute) Contributions of Variables to the Principal Components

```{r}
library("factoextra")
fviz_contrib(res.pca, choice = "var", axes = 1) + ggtitle("CTA Axis 1")
fviz_contrib(res.pca, choice = "var", axes = 2) + ggtitle("CTA Axis 2")
```
#### PCA Variables Output - 2 Axis

Making a variable table output for interpretation with Coordinates, CTA, CTR and CTR Sum of the principal plan. I use this output in a table form with all analysis together since it's easy to interpret and export to any document. 

CTA - (Absolute) Contributions of Variables to the Principal Components
CTR - (Relative) Contributions of the axis to the variables (Quality of Representation)


```{r}
vardf_2 <- data.frame(res.pca$var$coord[,1:2], res.pca$var$contrib[,1:2], res.pca$var$cos2[,1:2]*100)
colnames(vardf_2) <- c('PC1', 'PC2', 'CTA1', 'CTA2', 'CTR1', 'CTR2')

#reordering the variable dataframe
vardf_2 <- vardf_2[,c(1,3,5,2,4,6)]

#rounding to 2 decimal places
vardf_2 <- round(vardf_2,2)

#Adding CTR Sum for Quality of Representation
vardf_2[7] <- vardf_2[3] + vardf_2[6]
colnames(vardf_2)[colnames(vardf_2) == 'CTR1.1'] <- 'CTR Sum'
vardf_2

write.table(vardf_2, "d:/mydata.txt", sep="\t")

```

#### PCA Variables Output - 3 Axis

```{r}
vardf_3 <- data.frame(res.pca$var$coord[,1:3],res.pca$var$contrib[,1:3], res.pca$var$cos2[,1:3]*100)
colnames(vardf_3) <- c('PC1', 'PC2', 'PC3', 'CTA1', 'CTA2', 'CTA3', 'CTR1', 'CTR2', 'CTR3')

#reordering the variable dataframe
vardf_3 <- vardf_3[,c(1,4,7,2,5,8,3,6,9)]

#rounding to 2 decimal places
vardf_3 <- round(vardf_3,2)

#Adding CTR Sum for Quality of Representation
vardf_3[10] <- vardf_3[3] + vardf_3[6] + vardf_3[9]
colnames(vardf_3)[colnames(vardf_3) == 'CTR1.1'] <- 'CTR Sum'

vardf_3
```
#### Individuals PCA Analysis


CTA - (Absolute) Contributions of Individuals to the Principal Components


```{r}
fviz_contrib(res.pca, choice = "ind", axes = 1) + ggtitle("CTA Axis 1")
fviz_contrib(res.pca, choice = "ind", axes = 2) + ggtitle("CTA Axis 2")
```

#### PCA Individuals Output - 2 Axis

Making an individual table output for interpretation with Coordinates, CTA, CTR, CTR Sum of the principal plan

```{r}
inddf_2 <- data.frame(res.pca$ind$coord[,1:2],res.pca$ind$contrib[,1:2], res.pca$ind$cos2[,1:2]*100)
colnames(inddf_2) <- c('PC1', 'PC2', 'CTA1', 'CTA2', 'CTR1', 'CTR2')

#reordering the variable dataframe
inddf_2 <- inddf_2[,c(1,3,5,2,4,6)]

#rounding to 2 decimal places
inddf_2 <- round(inddf_2,2)

#Adding CTR Sum for Quality of Representation
inddf_2[7] <- inddf_2[3] + inddf_2[6]
colnames(inddf_2)[colnames(inddf_2) == 'CTR1.1'] <- 'CTR Sum'
rownames(inddf_2) <- df$Anos

inddf_2

```
#### PCA Individuals Output - 3 Axis

```{r}
inddf_3 <- data.frame(res.pca$ind$coord[,1:3],res.pca$ind$contrib[,1:3], res.pca$ind$cos2[,1:3]*100)
colnames(inddf_3) <- c('PC1', 'PC2', 'PC3', 'CTA1', 'CTA2', 'CTA3', 'CTR1', 'CTR2', 'CTR3')

#reordering the variable dataframe
inddf_3 <- inddf_3[,c(1,4,7,2,5,8,3,6,9)]

#rounding to 2 decimal places
inddf_3 <- round(inddf_3,2)

#Adding CTR Sum for Quality of Representation
inddf_3[10] <- inddf_3[3] + inddf_3[6] + inddf_3[9]
colnames(inddf_3)[colnames(inddf_3) == 'CTR1.1'] <- 'CTR Sum'
rownames(inddf_3) <- df$Anos

inddf_3
```
#### Correlation Circle

Correlation of variables with principal components 1 and 2 (Correlation Circle)

```{r}
res.desc <- dimdesc(res.pca, axes = 1:2, proba = 0.05)
res.desc$Dim.1
res.desc$Dim.2

```


